A SAT-Based Counterexample Guided Method for Unbounded Synthesis

### 2016-01-27 paper.tex
Bounded section from cav'14 paper

@@

We review the bounded synthesis algorithm by Narodytska et al.~\cite{narodytska2014},
which is the main building block for our unbounded algorithm.

A set of runs can be represented symbolically by a tree of valuations to
controllable and uncontrollable label variables where valuations are allowed to
be \emph{unfixed}. An unfixed value symbolically represents all possible
valuations. Formally, we define the tree as a set of lists of label valuations
and a node in the tree may be identified by the list made by following the path
from the root.

The entire set of runs of a bounded game of rank $n$ is symbolically
represented by a tree of depth $2n$ (a step is two moves - environment then
controller) and width 1 populated by unfixed edges.  Fixing some moves and
thereby reducing the set of runs forms an \emph{abstract game tree} that
symbolically represents a smaller version of the game.

A strategy is equivalent to the set of all runs with the controller's labels
obeying the strategy mapping $\pi$. Therefore, a strategy can also be
represented by a tree. Relaxing the restriction of the strategy mapping allows
for a \emph{partial strategy} in which multiple labels are now possible for a
single state.

A strategy or partial strategy can also be thought of as an abstract game. A
partial strategy for the controller is a restriction only on the controllable
labels in the game. So if the environment can not win in the abstract game
equivalent to the controller's partial strategy, then all strategies allowable
by that partial strategy must be winning.

An abstract game tree can be checked for the existence of a winning run by a
SAT solver\cite{narodytska2014}.  The tree must be encoded into CNF by making
copies of the transition relation for each game step in tree. The formula must
also check whether the error state has been reached in no branch
(\emph{respectively}: in all branches) for the controller (environment.) The
functions that produce these formulas are shown in Algorithm
\ref{alg:treeFormula}.

When we produce a formula using $\textsc{treeFormula}$ the SAT solver is
playing on behalf on the controller. Any unfixed labels in the tree,
controllable or uncontrollable, will be existentially quantified.  This means
that if there exists a way for the game to be solved when both players
cooperate on unfixed labels the SAT solver will find it. If no winning run
exists in an abstract game even when the players are cooperating then there is
no winning run when the opponent is playing adversarily.

By checking for the existence of a winning run for the environment in an
abstract game tree constructed from a partial controller strategy we are
checking if the partial strategy is winning. If no environment-winning run is
exists within the partial strategy then it is a winning partial strategy,
otherwise the spoiling run is a counterexample that proves that the partial
strategy is not winning. This forms the basis of a counterexample guided
abstraction refinement framework \cite{clarke2000} that operates on candidate strategies.

The bounded synthesis algorithm (Algorithm \ref{alg:bounded}) begins with an
empty game as seen in Figure \ref{fig:emptytree}.  Initially we are playing on
behalf of the environment because it chooses the first move in each step.  The
empty game is passed to the SAT solver, which searches for a candidate
environment strategy.  If a candidate is found (Figure \ref{fig:candidate})
then it is checked for a spoiling strategy by solving for the controller.  If
no spoiling strategy exists, that means our candidate is a winning strategy and
the algorithm terminates.  Otherwise we find a counterexample (Figure
\ref{fig:cex}), which is used to refine the empty game tree to include the
first move from the controller's spoiling strategy (Figure \ref{fig:refined}).
The algorithm continues by finding a new candidate for the environment (Figure
\ref{fig:candidate2}), and a new counterexample to refine the game abstraction
once again (Figure \ref{fig:refined2}). The full procedure is listed in
Algorithm \ref{alg:bounded}.

\tikzset{every node/.style={solid}}
\tikzstyle{fixed}=[solid]
\tikzstyle{unfixed}=[dash pattern = on 2pt off 2pt]
\begin{figure}
    \centering
    \begin{subfigure}{.3\textwidth}
        \centering
        \begin{tikzpicture}[dash pattern = on 2pt off 2pt, level distance = 10mm]
            \node [circle,draw] (root){}
                child {node [circle,draw] {} edge from parent[unfixed]
                    child {node [circle,draw] {} edge from parent[unfixed]}
                };
        \end{tikzpicture}
        \caption{An empty AGT}
        \label{fig:emptytree}
    \end{subfigure}%
    \begin{subfigure}{.35\textwidth}
        \centering
        \begin{tikzpicture}[dash pattern = on 2pt off 2pt, level distance = 10mm]
            \node [circle,draw] (root) {}
                child {node [circle,draw] {}
                    child {node [circle,draw] {} edge from parent[unfixed]}
                    edge from parent [fixed] node [left] {$u_1$}
                };
        \end{tikzpicture}
        \caption{Environment candidate}
        \label{fig:candidate}
    \end{subfigure}%
    \begin{subfigure}{.4\textwidth}
        \centering
        \begin{tikzpicture}[dash pattern = on 2pt off 2pt, level distance = 10mm]
            \node [circle,draw] (root){}
                child {node [circle,draw] {} edge from parent[unfixed]
                    child {node [circle,draw] {} edge from parent[fixed] node [left] {$c_1$}}
                    edge from parent [fixed] node [left] {$u_1$}
                };
        \end{tikzpicture}
        \caption{Counterexample}
        \label{fig:cex}
    \end{subfigure}

    \begin{subfigure}{.3\textwidth}
        \centering
        \begin{tikzpicture}[dash pattern = on 2pt off 2pt, level distance = 10mm]
            \node [circle,draw] (root){}
                child {node [circle,draw] {} edge from parent[unfixed]
                    child {node [circle,draw] {} edge from parent[fixed] node [left] {$c_1$}}
                };
        \end{tikzpicture}
        \caption{Refined AGT}
        \label{fig:refined}
    \end{subfigure}%
    \begin{subfigure}{.35\textwidth}
        \centering
        \begin{tikzpicture}[dash pattern = on 2pt off 2pt, level distance = 10mm]
            \node [circle,draw] (root) {}
                child {node [circle,draw] {}
                    child {node [circle,draw] {} edge from parent[unfixed]}
                    edge from parent [fixed] node [left] {$u_2$}
                };
        \end{tikzpicture}
        \caption{New candidate}
        \label{fig:candidate2}
    \end{subfigure}%
    \begin{subfigure}{.4\textwidth}
        \centering
        \begin{tikzpicture}[dash pattern = on 2pt off 2pt, level distance = 10mm]
            \node [circle,draw] (root){}
                child {node [circle,draw] {} edge from parent[unfixed]
                    child {node [circle,draw] {} edge from parent[fixed] node [left] {$c_1$}}
                    child {node [circle,draw] {} edge from parent[fixed] node [right] {$c_2$}}
                };
        \end{tikzpicture}
        \caption{Refined AGT}
        \label{fig:refined2}
    \end{subfigure}
    \caption{Abstract game trees}
\end{figure}

@@

>>>>

\subsection{Abstract Game Trees}

\tikzset{every node/.style={solid}}
\tikzstyle{fixed}=[solid]
\tikzstyle{unfixed}=[dash pattern = on 2pt off 2pt]
\begin{figure}
    \centering
    \begin{subfigure}[t]{.25\textwidth}
        \centering
        \begin{tikzpicture}[dash pattern = on 2pt off 2pt, level distance = 10mm]
            \node [circle,draw] (root){}
                child {node [circle,draw] {}
                    child {node [circle,draw] {}
                        edge from parent [fixed] node [left] {$b$}
                    }
                    edge from parent [fixed] node [left] {$a$}
                }
                child {node [circle,draw] {}
                    edge from parent [fixed] node [right] {$d$}
                }
                node [above=4pt] {$\langle s, k \rangle$};
        \end{tikzpicture}
        \captionsetup{width=.8\textwidth}
        \caption{AGT}
        \label{fig:agt}
    \end{subfigure}%
    \begin{subfigure}[t]{.25\textwidth}
        \centering
        \begin{tikzpicture}[dash pattern = on 2pt off 2pt, level distance = 10mm]
            \node [circle,draw] (root){}
                child {node [circle,draw] {}
                    child {node [circle,draw] {}
                        node [left=4pt] {$\epsilon$}
                        edge from parent [fixed] node [left] {$b$}
                    }
                    node [left=4pt] {$\beta$}
                    edge from parent [fixed] node [left] {$a$}
                }
                child {node [circle,draw] {}
                    node [left=4pt] {$\gamma$}
                    edge from parent [fixed] node [right] {$d$}
                }
                node [right=4pt] {$\alpha$}
                node [above=4pt] {$\langle s, k \rangle$};
        \end{tikzpicture}
        \captionsetup{width=.8\textwidth}
        \caption{AGT with strategy}
        \label{fig:strategy}
    \end{subfigure}%
    \begin{subfigure}[t]{.25\textwidth}
        \centering
        \begin{tikzpicture}[dash pattern = on 2pt off 2pt, level distance = 10mm]
            \node [circle,draw] (root){}
                child {node [circle,draw] {}
                    child {node [circle,draw] {}
                        node [left=4pt] {$\epsilon$}
                        edge from parent [fixed] node [left] {$b$}
                    }
                    node [left=4pt] {$\beta$}
                    edge from parent [fixed] node [left] {$a$}
                }
                child {node [circle,draw] (sroot) {}
                    child {node [below left=0 and -10pt] (sa) {}
                        edge from parent [draw=none]
                    }
                    child {node [below right=0 and -10pt] (sb) {}
                        edge from parent [draw=none]
                    }
                    node [left=4pt] {$\gamma$}
                    edge from parent [fixed] node [right] {$d$}
                }
                node [right=4pt] {$\alpha$}
                node [above=4pt] {$\langle s, k \rangle$};
                \draw[unfixed] (sa.center) -- (sb.center);
                \draw[unfixed] (sroot.south) -- (sa.center);
                \draw[unfixed] (sroot.south) -- (sb.center);
        \end{tikzpicture}
        \captionsetup{width=.8\textwidth}
        \caption{AGT with spoiling strategy}
        \label{fig:spoiling}
    \end{subfigure}%
    \begin{subfigure}[t]{.25\textwidth}
        \centering
        \begin{tikzpicture}[dash pattern = on 2pt off 2pt, level distance = 10mm]
            \node [circle,draw] (root){}
                child {node [circle,draw] {}
                    child {node [circle,draw] {}
                        edge from parent [fixed] node [left] {$b$}
                    }
                    edge from parent [fixed] node [left] {$a$}
                }
                child {node [circle,draw] {}
                    child {node [circle,draw] {}
                        edge from parent [fixed] node [right] {$e$}
                    }
                    edge from parent [fixed] node [right] {$d$}
                }
                node [above=4pt] {$\langle s, k \rangle$};
        \end{tikzpicture}
        \captionsetup{width=.8\textwidth}
        \caption{Refined AGT}
        \label{fig:refined}
    \end{subfigure}
%%%    \begin{subfigure}{.3\textwidth}
%%%        \centering
%%%        \begin{tikzpicture}[dash pattern = on 2pt off 2pt, level distance = 10mm]
%%%            \node [circle,draw] (root){}
%%%                child {node [circle,draw] {} edge from parent[unfixed]
%%%                    child {node [circle,draw] {} edge from parent[unfixed]}
%%%                };
%%%        \end{tikzpicture}
%%%        \caption{An empty AGT}
%%%        \label{fig:emptytree}
%%%    \end{subfigure}%
%%%    \begin{subfigure}{.35\textwidth}
%%%        \centering
%%%        \begin{tikzpicture}[dash pattern = on 2pt off 2pt, level distance = 10mm]
%%%            \node [circle,draw] (root) {}
%%%                child {node [circle,draw] {}
%%%                    child {node [circle,draw] {} edge from parent[unfixed]}
%%%                    edge from parent [fixed] node [left] {$u_1$}
%%%                };
%%%        \end{tikzpicture}
%%%        \caption{Environment candidate}
%%%        \label{fig:candidate}
%%%    \end{subfigure}

%%%    \begin{subfigure}{.4\textwidth}
%%%        \centering
%%%        \begin{tikzpicture}[dash pattern = on 2pt off 2pt, level distance = 10mm]
%%%            \node [circle,draw] (root){}
%%%                child {node [circle,draw] {} edge from parent[unfixed]
%%%                    child {node [circle,draw] {} edge from parent[fixed] node [left] {$c_1$}}
%%%                    edge from parent [fixed] node [left] {$u_1$}
%%%                };
%%%        \end{tikzpicture}
%%%        \caption{Counterexample}
%%%        \label{fig:cex}
%%%    \end{subfigure}%
%%%    \begin{subfigure}{.3\textwidth}
%%%        \centering
%%%        \begin{tikzpicture}[dash pattern = on 2pt off 2pt, level distance = 10mm]
%%%            \node [circle,draw] (root){}
%%%                child {node [circle,draw] {} edge from parent[unfixed]
%%%                    child {node [circle,draw] {} edge from parent[fixed] node [left] {$c_1$}}
%%%                };
%%%        \end{tikzpicture}
%%%        \caption{Refined AGT}
%%%        \label{fig:refined}
%%%    \end{subfigure}%
%%%    \begin{subfigure}{.35\textwidth}
%%%        \centering
%%%        \begin{tikzpicture}[dash pattern = on 2pt off 2pt, level distance = 10mm]
%%%            \node [circle,draw] (root) {}
%%%                child {node [circle,draw] {}
%%%                    child {node [circle,draw] {} edge from parent[unfixed]}
%%%                    edge from parent [fixed] node [left] {$u_2$}
%%%                };
%%%        \end{tikzpicture}
%%%        \caption{New candidate}
%%%        \label{fig:candidate2}
%%%    \end{subfigure}%
%%%    \begin{subfigure}{.4\textwidth}
%%%        \centering
%%%        \begin{tikzpicture}[dash pattern = on 2pt off 2pt, level distance = 10mm]
%%%            \node [circle,draw] (root){}
%%%                child {node [circle,draw] {} edge from parent[unfixed]
%%%                    child {node [circle,draw] {} edge from parent[fixed] node [left] {$c_1$}}
%%%                    child {node [circle,draw] {} edge from parent[fixed] node [right] {$c_2$}}
%%%                };
%%%        \end{tikzpicture}
%%%        \caption{Refined AGT}
%%%        \label{fig:refined2}
%%%    \end{subfigure}
    \caption{Abstract game trees}
    \label{fig:alltrees}
\end{figure}

We review the bounded synthesis algorithm by Narodytska et
al.~\cite{narodytska2014}, which is the main building block for our unbounded
algorithm.  The algorithm constructs a series of abstractions of the input
game.  An abstraction restricts actions available to one of the players.
Specifically, we consider abstractions represented as trees of actions,
referred to as \emph{abstract game trees}.  Together with a state-round pair
$\langle s,k\rangle$, an abstract game tree defines an \emph{abstract game}
played from this state.  Figure~\ref{fig:agt} shows an example abstract game
tree.  In the abstract game, the environment player is required to pick actions
from the tree, starting from the root node.  After reaching a leaf, it
continues playing unrestricted.  The tree in Figure~\ref{fig:agt} restricts the
initial environment action to the set $\{a, d\}$.  After choosing action $d$,
the environment reaches a leaf of the tree and continues playing unrestricted.
Alternatively, after choosing $a$, the environment is required to play action
$b$ in the next round.

Nodes of an abstract game tree are uniquely identified by the list of edge
labels along the path from the root to the node.  We identify an abstract game
tree with the set of its nodes.  For example, the tree in Figure~\ref{fig:agt}
can be written as $\{(),(d),(a),(a,b)\}$.  We denote $\textsc{leaves}(T)$ the
subset of leaf nodes of a tree $T$.

A \emph{partial strategy} $Strat: T \rightarrow L_c $ assigns a controllable
action to be played in each node of the abstract game tree.
Figure~\ref{fig:strategy} shows an example partial strategy.  The controller
starts by choosing action $\alpha$.  If the environment plays $a$, the
controller responds with $\beta$ in the next round, and so on. Given a partial
strategy $Strat$, we can map each leaf $l$ of the abstract game tree to
$\langle s',i'\rangle=\textsc{outcome}(\langle s, i\rangle, Strat, l)$ obtained
by playing all controllable and uncontrollable actions on the path from the
root to the leaf.

%%%A set of runs can be represented symbolically by a tree of valuations to
%%%controllable and uncontrollable label variables where valuations are allowed to
%%%be \emph{unfixed}. An unfixed value symbolically represents all possible
%%%valuations. Formally, we define the tree as a set of lists of label valuations
%%%and a node in the tree may be identified by the list made by following the path
%%%from the root.

%%%The entire set of runs of a bounded game of rank $n$ is symbolically
%%%represented by a tree of depth $2n$ (a step is two moves - environment then
%%%controller) and width 1 populated by unfixed edges.  Fixing some moves and
%%%thereby reducing the set of runs forms an \emph{abstract game tree} that
%%%symbolically represents a smaller version of the game.

%%%A strategy is equivalent to the set of all runs with the controller's labels
%%%obeying the strategy mapping $\pi$. Therefore, a strategy can also be
%%%represented by a tree. Relaxing the restriction of the strategy mapping allows
%%%for a \emph{partial strategy} in which multiple labels are now possible for a
%%%single state.

%%%A strategy or partial strategy can also be thought of as an abstract game. A
%%%partial strategy for the controller is a restriction only on the controllable
%%%labels in the game. So if the environment can not win in the abstract game
%%%equivalent to the controller's partial strategy, then all strategies allowable
%%%by that partial strategy must be winning.

%%%An abstract game tree can be checked for the existence of a winning run by a
%%%SAT solver\cite{narodytska2014}.  The tree must be encoded into CNF by making
%%%copies of the transition relation for each game step in tree. The formula must
%%%also check whether the error state has been reached in no branch
%%%(\emph{respectively}: in all branches) for the controller (environment.) The
%%%functions that produce these formulas are shown in Algorithm
%%%\ref{alg:treeFormula}.

%%%When we produce a formula using $\textsc{treeFormula}$ the SAT solver is
%%%playing on behalf on the controller. Any unfixed labels in the tree,
%%%controllable or uncontrollable, will be existentially quantified.  This means
%%%that if there exists a way for the game to be solved when both players
%%%cooperate on unfixed labels the SAT solver will find it. If no winning run
%%%exists in an abstract game even when the players are cooperating then there is
%%%no winning run when the opponent is playing adversarily.

%%%By checking for the existence of a winning run for the environment in an
%%%abstract game tree constructed from a partial controller strategy we are
%%%checking if the partial strategy is winning. If no environment-winning run
%%%exists within the partial strategy then it is a winning partial strategy,
%%%otherwise the spoiling run is a counterexample that proves that the partial
%%%strategy is not winning. This forms the basis of a counterexample guided
%%%abstraction refinement framework \cite{clarke2000} that operates on candidate strategies.

@@

@@

Algorithm~\ref{alg:bounded} illustrates the bounded synthesis algorithm. The
algorithm takes a concrete game $G$ as an implicit argument. In addition, it
takes a player (controller or environment), state-round pair $\langle s,
k\rangle$ and an abstract game tree $T$ and returns a winning partial strategy
for it, if one exists.  The initial invocation of the algorithm takes the
initial state $\langle I, 0\rangle$ and an empty abstract game tree
$\emptyset$. Initially the solver is playing on behalf of the environment since
that player takes the first move in every step.  The empty game tree does not
constrain opponent moves, hence solving such an abstraction is equivalent to
solving the original concrete game.  The algorithm is organised as a
counterexample-guided abstraction refinement (CEGAR) loop.  The first step of
the algorithm uses the \textsc{findCandidate} function, described in detail
below, to come up with a candidate partial strategy for $T$. If it fails to
find a strategy, this means that no winning partial strategy exists for $T$.
If, on the other hand, a candidate partial strategy is found, we need to verify
if it is indeed winning for $T$.

The \textsc{verify} procedure searches for a \emph{spoiling} counterexample
strategy in each leaf of the candidate partial strategy by calling
\textsc{solveAbstract} for the opponent. The dual solver solves games on behalf
of the controller player.  {Figure~\ref{fig:spoiling}} shows a spoiling
strategy discovered one of the leaves of the partial strategy.

If the dual solver can find no spoiling strategy at any of the leaves, then the
candidate partial strategy is a winning one. Otherwise, \textsc{verify} returns
the move used by the opponent to defeat a leaf of the partial strategy, which
is appended to the corresponding node in $T$ in order to refine it in line~(9)
as shown in Figure~(\ref{fig:refined}).

We solve the refined game by recursively invoking \textsc{solveAbstract} on it.
If no partial winning strategy is found for the refined game then there is also
no partial winning strategy for the original abstract game, and the algorithm
returns a failure.  Otherwise, the partial strategy for the refined game is
\emph{projected} on the original abstract game by removing the leaves
introduced by refinements. The resulting partial strategy becomes a candidate
strategy to be verified at the next iteration of the loop. In the worst case
the loop terminates after all actions in the game are refined into the abstract
game.

The way game trees are refined is determined by the formula passed to the SAT
solver in $\textsc{findCandidate}$. This formula encodes an abstract game tree
into CNF by making copies of the transition relation for each game step in the
tree.  The formula must also check whether the error state has been reached in
no branch (\emph{respectively}: in all branches) for the controller
(environment.) This procedure is defined in Algorithm \ref{alg:treeFormula} for
both the controller in $\textsc{treeFormula}$ and the environment in
$\textsc{\textoverline{treeFormula}}$.

Since the game tree formulation is passed to a SAT solver, any unfixed labels
in the tree - both controllable and uncontrollable - will be existentially
quantified. This means that the SAT solver will find any way to win the game
while both players are cooperating. If no winning run exists in an abstract
game even when the players are cooperating then there is no winning run when
the opponent is playing adversarily. When a winning run is found, the actions
chosen by the SAT solver are used to refine the game tree. This is advantageous
for many synthesis problems where the game must be formalised as adversarial
for correctness but the final implementation will be cooperating with its
environment in the real world. An example of such a system is a device driver
that cooperates with the device and OS to interface between the two.

%%%The bounded synthesis algorithm (Algorithm \ref{alg:bounded}) begins with an
%%%empty game as seen in Figure \ref{fig:emptytree}.  Initially we are playing on
%%%behalf of the environment because it chooses the first move in each step.  The
%%%empty game is passed to the SAT solver, which searches for a candidate
%%%environment strategy.  If a candidate is found (Figure \ref{fig:candidate})
%%%then it is checked for a spoiling strategy by solving for the controller.  If
%%%no spoiling strategy exists, that means our candidate is a winning strategy and
%%%the algorithm terminates.  Otherwise we find a counterexample (Figure
%%%\ref{fig:cex}), which is used to refine the empty game tree to include the
%%%first move from the controller's spoiling strategy (Figure \ref{fig:refined}).
%%%The algorithm continues by finding a new candidate for the environment (Figure
%%%\ref{fig:candidate2}), and a new counterexample to refine the game abstraction
%%%once again (Figure \ref{fig:refined2}). The full procedure is listed in
%%%Algorithm \ref{alg:bounded}.
